FYS3150 Computational Physics, project 1

### 2018-09-11 Project1/report/ms.tex

\documentclass{emulateapj}
%\documentclass[12pt,preprint]{aastex}

@@

\title{Project 1}

@@

\begin{abstract}
Comparing general and tailored methods for solving second order differential equations this study will present that producing a specialized method reduces the run time of a numerical solver without losing numerical accuracy. Reducing the floating point operations (FLOPS) will be proven efficient for large scale calculations.

  %State problem. Briefly describe method and data. Summarize main results.
\end{abstract}
\keywords{computational science --- numerical methods: error estimation, run time --- methods: numerical, tridiagonal matrices}

\section{Introduction}
\label{sec:introduction}
This study on generalized and tailored numerical algorithms will focus on how a differential equation is solved numerically most efficiently while not reducing the accuracy of the calculations. By comparing the run time and observing the numerical error of the calculations . To specify, the equation in question is the one dimensional Poisson equation:
%
\begin{equation*}
    {\nabla^{2}} \Phi = -4\pi \rho(r),
\end{equation*}
%
where $\Phi$ is the electrostatic potential generated by a localized charge distribution and $\rho$ is the charge density. This equation is used in electromagnetism, but in this case it was only necessary to analyze a normalized equation on the same form. Our normalized one dimensional Poisson equation can then be written as:
%
\begin{equation}\label{eq:-u''}
    -\frac{du^2}{dx^2} = f(x).
\end{equation}

Further rewriting this as a set of linear equation will allow us to solve it using our numerical algorithms. As generalized algorithm we will be using a pre-built function for LU-decomposition. Specializing our algorithm further for the problem we will make a generalized algorithm for any solving any tridiagonal matrix. Lastly we will make an algorithm completely tailored to our problem where the matrix elements are already specified.

@@

Having reduced the FLOPS to minimize the CPU time the calculations are timed to compare with the general algorithm. Timing is done with a built in function within C++ and only ran over the calculations of $\boldsymbol{A}\boldsymbol{u} = \tilde{f}$.

@@

\subsection{LU-decomposition}
Comparing the results with a completely general method we solve $\boldsymbol{A}\boldsymbol{u} = \boldsymbol{f}$ with by LU decomposition with armadillo using a solve function.

For a non-singular matrix $\boldsymbol{A} \in \R ^{nxn}$ we can also preform LU-decomposition, given by:
%
\[ \boldsymbol{A} = \boldsymbol{Lu} =
\begin{array}{c}
\begin{bmatrix}\label{eq:A = Lu}
l_{11}     & 0           & \dots         & 0       & 0 \\
l_{21}     & l_{22}           & 0       & \dots      & 0       \\
\dots       & \dots           & \dots       & \dots      & \dots    \\
\dots       & \dots           & \dots       & \dots      & 0    \\
 l_{n1} &  l_{n2}            & \dots        & l_{nn-1}      & l_{nn}     \\
\end{bmatrix}
\begin{bmatrix}
u_{11}     & u_{12}            & \dots         & \dots       & u_{1n} \\
0     & u_{22}           & \dots       & \dots      & u_{2n}       \\
\dots       & \dots           & \dots       & \dots      & \dots    \\
0       & \dots           & \dots       & \dots      & u_{(n-1)n}    \\
 0 &  0            & \dots        & 0      & u_{nn}     \\
\end{bmatrix}

@@

%Summarize properties of data. Which data are used (experiment,frequencies etc.)? Pixel resolution ($N_{\textrm{side}}$),$\ell_{\textrm{max}}$ -- everything necessary to repeat the analysis for other researchers.

@@

\begin{figure}[t]
\mbox{\epsfig{figure=plot1d.png,width=\linewidth,clip=}}
\caption{Logarithmic plot of the error as function of the step size, to illustrate the $\mathcal{O}(h^{2})$ dependency we have plotted $log_{10}(h^{2})=2log_{10}(h)$, which in the plot doubles as $y = 2x$, the expected value of our error.}
\label{fig:1d}
\end{figure}
%
\begin{deluxetable}{lcc}[H]
%\tablewidth{0pt}
\tablecaption{\label{tab:results}}
\tablecomments{Table of the error, for stepsizes down to $h = 10^{-7}$. The table reveals that the numerical error starts increasing after $h = 10^{-5}$, and before reaching that point we the the $\mathcal{O}(h^{2})$.}
\tablecolumns{4}
\tablehead{ Step size & Error &}
\startdata
 \\
 $10^{-1}$ & $6.6114*10^{-2}$ &\\
 $10^{-2}$ & $8.1651*10^{-4}$ &\\
 $10^{-3}$ & $8.3167*10^{-6}$ &\\
 $10^{-4}$ & $8.3311*10^{-8}$ &\\
 $10^{-5}$ & $1.4356*10^{-9}$ &\\
 $10^{-6}$ & $8.3973*10^{-7}$ &\\
 $10^{-7}$ & $2.7487*10^{-6}$ &\\
\enddata
\end{deluxetable}

@@

Testing the time we ran the three different programs 10 times for different sizes $n$, listing longest, shortest and average CPU times.
In [TABLE ] the results we have listed for timing our programs with $n = 1000$. Comparably for larger matrices $n=10^{6}$ the programs yield the timings presented in [TABLE ].
%
\begin{deluxetable}{lccc}[H]
%\tablewidth{0pt}
\tablecaption{\label{tab:results}}
\tablecomments{This table contains highest, lowest and average measured CPU times for the different algorithms used at n = 1000, armadillo being overall, while our general and tailored algorithm timings both withing the same order of magnitude.}
\tablecolumns{4}
\tablehead{ & General & Tailored & Armadillo}
\startdata
Highest & 10.8*10^{-5}s & 8.4*10^{-5}s & 0.012942s \\
Lowest & 3.4*10^{-5}s & 2.5*10^{-5}s & 0.007715s  \\
Average & 7.8*10^{-5}s & 5.5*10^{-5}s & 0.010166s 
\enddata
\end{deluxetable}
%
%
\begin{deluxetable}{lccc}[h]
%\tablewidth{0pt}
\tablecaption{\label{tab:results}}
\tablecomments{This table contains highest, lowest and average measured CPU times for the different algorithms used at $n = 10^{6}$, armadillo runs out of memory, while the general and tailored algorithms staying within the same order of magnitude.}
\tablecolumns{4}
\tablehead{ & General & Tailored & Armadillo}
\startdata
Highest & 0.029865s & 0.028870s & Out of memory \\
Lowest & 0.027048s & 0.027058s & Out of memory  \\
Average & 0.027544s & 0.027602s & Out of memory 
\enddata
\end{deluxetable}
%

Note that the LU-decomposition runs out of memory at $n = 10^{5}$ meaning that we will not be able to test the Armadillo made algorithm for larger matrices. While the tailored and general algorithms are of same order of magnitude and even similar in size (on average) for $n = 10^{7}$, we see a clear difference for the smaller $n = 1000$. Our general algorithm runs at $8n$ FLOPS, while our specialized algorithm only runs at $6n$ FLOPS.

@@

@@

\bibitem[G{\'o}rski et al.(1994)]{gorski:1994} G{\'o}rski, K. M.,
  Hinshaw, G., Banday, A. J., Bennett, C. L., Wright, E. L., Kogut,
  A., Smoot, G. F., and Lubin, P.\ 1994, ApJL, 430, 89

>>>>

\documentclass{emulateapj}
\documentclass[12pt,preprint]{aastex}

@@

\title{FYS3150 Computational Physics, project 1}

@@

\begin{abstract}
Comparing methods for solving second order differential equations this study will present that producing a specialized method reduces the run time of a numerical solver. When solving a second order differential equation we find that tailoring a method to the problem at hand will result in a shorter CPU time compared to a general method similar to the specialized. Further comparison proves that even slightly tailoring the algorithm (by the general method) will prove to be faster than a standard LU decomposition. For large scale calculations the LU decomposition is found to use more memory than the computer has available, while the general and specialized methods produces results, the latter being faster. The specialized method will focus on reducing the floating point operations (FLOPS) by tailoring the equations prior to the numerical calculations.

  %State problem. Briefly describe method and data. Summarize main results.
\end{abstract}
\keywords{computational science --- numerical methods: error estimation, LU decomposition, forward/backward substitution --- methods: analytical, numerical, statistical}

\section{Introduction}
\label{sec:introduction}
This study on generalized and specialized numerical algorithms will focus on how a differential equation is solved numerically most efficiently while not reducing the accuracy of the calculations. By comparing the run time and observing the numerical error of the calculations . To specify, the equation in question is the one dimensional Poisson equation:
%
\begin{equation*}
    {\nabla^{2}} \Phi = -4\pi \rho(r),
\end{equation*}
%
where $\Phi$ is the electrostatic potential generated by a localized charge distribution and $\rho$ is the charge density. This equation is used in electromagnetism, but in this case it was only necessary to analyze a normalized equation on the same form. Our normalized one dimensional Poisson equation can then be written as:
%
\begin{equation}\label{eq:-u''}
    -\frac{du^2}{dx^2} = f(x).
\end{equation}

Further rewriting this as a set of linear equation will allow us to solve it using our numerical algorithms. As generalized algorithm we will be using a pre-built function for LU-decomposition. Specializing our algorithm further for the problem we will make a generalized algorithm for any solving any tridiagonal matrix. Lastly we will make an algorithm completely specialized to our problem where the matrix elements are already specified.

@@

Having reduced the FLOPS to minimize the CPU time the calculations are timed to compare with the general algorithm. Timing is done with a built in function within C++ and only ran over the calculations of $\boldsymbol{A}\boldsymbol{u} = \boldsymbol{\tilde{f}}$. Pseudo code is presented below

@@

\subsection{LU-decomposition}
Using a package from
%
\[ \boldsymbol{A} = \boldsymbol{Lu} =
\begin{array}{c}
\begin{bmatrix}\label{eq:A = Lu}
l_{11}     & 0           & \dots         & 0       & 0 \\
l_{21}     & l_{22}           & 0       & \dots      & 0       \\
\dots       & \dots           & \dots       & \dots      & \dots    \\
\dots       & \dots           & \dots       & \dots      & 0    \\
 l_{n1} &  l_{n2}            & \dots        & l_{nn-1}      & l_{nn}     \\
\end{bmatrix}
\begin{bmatrix}
u_{11}     & u_{12}            & \dots         & \dots       & u_{1n} \\
0     & u_{22}           & \dots       & \dots      & u_{2n}       \\
\dots       & \dots           & \dots       & \dots      & \dots    \\
0       & \dots           & \dots       & \dots      & u_{(n-1)n}    \\
 0 &  0            & \dots        & 0      & u_{nn}     \\
\end{bmatrix}

@@

@@

\begin{figure}[t]
\mbox{\epsfig{figure=plot1d.png,width=\linewidth,clip=}}
\caption{Logarithmic plot of the error as function of the step size, to illustrate the $\mathcal{O}(h^{2})$ dependency we have plotted $log_{10}(h^{2})=2log_{10}(h)$, which in the plot doubles as $y = 2x$, the expected value of our error.}
\label{fig:1d}
\end{figure}
%
\begin{deluxetable}{lcc}[H]
%\tablewidth{0pt}
\tablecaption{\label{tab:results}}
\tablecomments{Table of the error, for stepsizes down to $h = 10^{-7}$. The table reveals that the numerical error starts increasing after $h = 10^{-5}$, and before reaching that point we the the $\mathcal{O}(h^{2})$.}
\tablecolumns{4}
\tablehead{ n & Error &}
\startdata
 \\
 $10^{1}$ & $6.6114 \cdot 10^{-2}$ &\\
 $10^{2}$ & $8.1651 \cdot 10^{-4}$ &\\
 $10^{3}$ & $8.3167\cdot 10^{-6}$ &\\
 $10^{4}$ & $8.3311\cdot 10^{-8}$ &\\
 $10^{5}$ & $1.4356\cdot 10^{-9}$ &\\
 $10^{6}$ & $8.3973\cdot 10^{-7}$ &\\
 $10^{7}$ & $2.7487\cdot 10^{-6}$ &\\
\enddata
\end{deluxetable}

@@

Testing the time we ran the three different programs 10 times for different sizes $n$, listing longest, shortest and average CPU times.
In [TABLE ] the results we have listed for timing our programs with $n = 1000$. Comparably for larger matrices $n=10^{6}$ the programs yield the timings presented in [TABLE ].
%
\begin{deluxetable}{lccc}[H]
%\tablewidth{0pt}
\tablecaption{\label{tab:results}}
\tablecomments{This table contains highest, lowest and average measured CPU times for the different algorithms used at n = 1000, armadillo being overall, while our general and specialized algorithm timings both withing the same order of magnitude.}
\tablecolumns{4}
\tablehead{ & General & Specialized & Armadillo}
\startdata
Highest & 0.000333s & 0.000237s & 0.021747s \\
Lowest & 0.000191s & 0.000074s & 0.012255s  \\
Average & 0.000249s & 0.000167s & 0.015978s 
\enddata
\end{deluxetable}
%
%
\begin{deluxetable}{lccc}[h]
%\tablewidth{0pt}
\tablecaption{\label{tab:results}}
\tablecomments{This table contains highest, lowest and average measured CPU times for the different algorithms used at $n = 10^{6}$, armadillo runs out of memory, while the general and specialized algorithms staying within the same order of magnitude.}
\tablecolumns{4}
\tablehead{ & General & Specialized & Armadillo}
\startdata
Highest & 0.085026 & 0.073819s & Out of memory \\
Lowest & 0.074668 & 0.064465s & Out of memory  \\
Average & 0.079288s & 0.067805s & Out of memory 
\enddata
\end{deluxetable}
%

Note that the LU-decomposition runs out of memory at $n = 10^{5}$ meaning that we will not be able to test the Armadillo made algorithm for larger matrices. Our general algorithm runs at $8n$ FLOPS, while our specialized algorithm is optimized to only run at $6n$ FLOPS.

@@

Reviewing the plots from [FIG:1], there is an eye catching difference in error made between $n = 10$ and $n = 100$ points. Which is expected of our algorithm considering the quadratic dependency of the error with respect to step size $h^{2}$. However between these step sizes we already reach the maximum resolution of a side by side comparison, and between $n = 100$ and $n = 1000$ the only visible change is in the resolution of the plot itself. aiding our understanding of the results further by computing the relative error and plotting it in a logarithmic scale we can see that our error indeed has a quadratic dependency. For larger step sizes however the computer starts introducing its own error

While the specialized and general algorithms are of same order of magnitude and though still not similar in size (on average) for $n = 10^{6}$, we see a clear difference for the smaller $n = 1000$. Analyzing the number of FLOPS in the different algorithms, it becomes clear that the specialized algorithm should be $8n/6n = 3/2$ times faster. Comparing average CPU times we see that this is highly accurate for $n=1000$, in which case its $0.000249/0.000167 = 1.49$times faster. However for $n = 10^{6}$ this does not seem to be the case entirely $0.079288/0.067805 = 1.17$times faster. This can partially be attributed to the small sample size of ten

@@

\bibitem{1}
  M. H. Jensen,
  \textit{Project 1},
  \url{https://github.com/CompPhysics/ComputationalPhysics/blob/master/doc/Projects/2018/Project1/pdf/Project1.pdf} (10.09.18)
